---
title: "Main functions for DIY R package"
author: "Fonti Kar"
date: "19/07/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE, 
                      eval = FALSE)

library(dplyr)
```

## DIY R package 

Below are some of the main functions you will need to create your first R package!

### The mise en place 

```{r Prep}
library(devtools) #Will load usesthis and roxgen2 as well
library(testthat)
```

### The main set up

```{r Set up first}
#Create the basic skeleton of your package
#create_package("package_name")

create_package("ohwhaley")
```

This function will create the basic skeleton of your R package

`.Rbuildignore` lists files that we need to have around but that should not be included when building the R package from source. More in 4.3.1.
`.Rproj.user`, if you have it, is a directory used internally by RStudio.
.gitignore anticipates Git usage and ignores some standard, behind-the-scenes files created by R and RStudio. Even if you do not plan to use Git, this is harmless.
`DESCRIPTION` provides metadata about your package. We edit this shortly.
`NAMESPACE` declares the functions your package exports for external use and the external functions your package imports from other packages. At this point, it is empty, except for a comment declaring that this is a file we will not edit by hand.
The `R/` directory is the “business end” of your package. It will soon contain .R files with function definitions.

**Optional but recommended** 

You can create an R package without using git, however one benefit of using git and GitHub is that you can then host your package in a Github repo. This allows you to easily share with others. The README.md will also auto-render so it looks pretty swish! There are many ways to use git with your existing R package or project - see [here](https://happygitwithr.com/new-github-first.html) for more info 

```{r Use git okay?}
#Initiate git and create a Github repo from R
#Assumes you have a personal access token (PAT) set up already
use_git() 
use_github()

#If you don't have a PAT set up yet
gh_token_help() #will guide you through troubleshoot and set up
create_github_token()
gitcreds::gitcreds_set()
```

### Edit your DESCRIPTION file manually

The DESCRIPTION file provides metadata about your package.

- Make yourself the author.
- Write some descriptive text in the `Title` and `Description` fields.

The `DESCRIPTION` file is where you specify any dependencies your package may have. You simply add the package name under `Depends/Imports/Suggests`.
If you want to use functions from existing packages, generally you will import most of your functions from other packages by using the `::` e.g `dplyr::mutate` and then adding dplyr under `Imports`

### Declare your license 

Read more about the different licenses in the [R packages book](https://r-pkgs.org/license.html)
```{r License up}
use_mit_license("Your name")
```

### Routine checks 

In order to make sure your R package can build and run smoothly, we will need regulary run some sanity checks. The goal is to get 0 WARNINGS, NOTES or ERRORS. The function is pretty clever and will provide you with google-able error messages if you do run into some trouble.

```{r Check it!}
check() 
```

### Your first function

```{r First function}
#Creates an R script called say.R in R/
use_r("say") 
use_r("phrases") 

#Copy and paste the contents of say.R and phrases.R in your empty R scripts
load_all() #Loads a preview of your early stage package

#Try use the function!
say()
```

🌟 EXTRA: 

- Want to write functions like dplyr where you can just type out the variable name instead of using data$var? You need to use curly braces in your function. For more details see https://dplyr.tidyverse.org/articles/programming.html 

```{r Data masking}
var_summary <- function(data, var) {
  data %>%
    summarise(n = n(), min = min({{ var }}), max = max({{ var }}))
}

mtcars %>% 
  group_by(cyl) %>% 
  var_summary(mpg)
```

### Document it! (Object documentation)

We will use roxygen tags to document our function. This will generate a helpfile for the `say()` function so when you type `?say()` into the console. This form of documentation is written in the same R script as your function. You can either copy the contents of the `roxygen_template.R` and paste above the `say()` function your `say.R`. 

OR

Place your cursor *inside* the `say()` function and click `Code > Insert Roxygen Skeleton` in RStudio.

Once you have filled out the roxygen tags with information relating to your function

```{r}
#Write documentation
document() #The function will go through your .R files and look at the roxygen tags and build/update any help files

#Notice that man/ has been added in your R package
#Notice that in NAMESPACE you can see say() because we used @export in the say() function

#Load your recent changes 
load_all() 

#Preview your recent changes
?say()
```

💡 TIPS: 

- Once you have a roxygen skeleton, hitting ENTER will continue the formatting on the next line `#'`
- Typing `@` and then hitting TAB will show you the various roxygen tags you can use for your documentation
- Use `@export` if you want to make your function external use, meaning anyone that downloads your package can use it!
- Use `@importFrom` to bring in LOTS of functions from other packages - `ggplot2` is a good example

```{r ggplot2}
#Instead of using the :: notation like below
my_plot <- function(data, x, y){
  ggplot2::ggplot(data, ggplot2::aes(x = x, y = y)) +
    ggplot2::ggplotgeom_point()
}

#You can import the functions using the roxygen tag as below
#' @importFrom ggplot2 ggplot aes geom_point theme scale_x_discrete     

```

### README - Your landing page

The function of your `README` is to give users a quick overview of what your R package is for and to help them get started with installation. Most R packages will have a installation instructions and some demo of its main functions. For this reason, its much easier to write your README in .Rmd format. Remember to knit regularly to preview your changes. 

Knitting your .Rmd will produce a .md. If you are using git and your package is hosted on GitHub, knit your .Rmd and commit and push both files. That way your README is updated on the repository

```{r README}
#This function will create the README file for you. 
#Copy and paste the contents of example_README_vign.Rmd into your empty file
use_readme_RMD()

#If your README is more basic and doesn't contain any code, just use the basic .md format.
use_readme_md()
```

🌟 EXTRA: 

- You can use GitHub Actions to auto-render your .md from .RMD - see https://fromthebottomoftheheap.net/2020/04/30/rendering-your-readme-with-github-actions/ 
In my experience so far... if your README is not very extensive and doesn't get updates often then the time used to set up GitHub Actions is not worth the trouble! 

### Long form documentation

Many great packages include helpful vignettes to so users can hit the ground running. Take a peak at `vignette("dplyr")`. Vignettes are usually a more extensive guide to using the package's functions with example data. Some parts may repeat the README. Try write your vignette with a beginner's mindset and explain things step by step.

You can create multiple vignettes too, perhaps include some background information or a contributing guideline so others can help add value to your package 👨‍💻🧑‍💻👩‍💻

```{r Vignette}
# use_vignette("packagename") #Creates .Rmd file for vignette inside vignette/
use_vignette("ohwhaley") 

# Copy and paste the contents of example_vignette.Rmd into your empty .Rmd file
# *For teaching purposes sake this is just a duplicate of the README

build_vignettes() #Build the vignette

#browseVignettes("packagename") #Previews in web browser
browseVignettes("ohwhaley") 

#vignette("packagename") #Previews in help window
vignette("ohwhaley") 
```

### Including data in your package

Some packages have built-in data that is available for the user. In some cases, these can be used for demonstration purposes e.g. the `starwars` dataset in `dplyr`. Alternatively, some packages that are compilations of datasets e.g https://github.com/thebioengineer/tidytuesdayR

```{r}
# Lets download some cetacean data from Tidy Tuesday
cetacean_subset <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-12-18/allCetaceanData.csv") %>% dplyr::select(species:birthYear)

# use_data(data_obj, name) #Saves the data as a .rda object in a data/ folder
use_data(cetacean_subset, name = cetacean)
```

Similar to a function, you will need to document your dataset so the meta-data can be found more easily like `?starwars`. To document the data, create a new .R script and save it as the name of the `.rda` object e.g. `cetacean.R` and store this with your functions in `R/`. 

There is no template on how to document data as far as I know but I have provided you with one in `data_template.R`

Copy and paste in the contents of `data_template.R` into  `cetacean.R`. Then run: 

```{r}
#To generate the helpfile
document() 

#Load recent changes
load_all()

#Preview your recent changes
?cetacean 
```

### Finally, lets talk about tests...


